Benchmarking R packages for calculation of persistent homology

### 2021-01-05 tdabench-paper/review-response.txt

        ---------- Reviewer ----------

        ## Overview ##

        I cannot comment a great deal on the overall suitability to the R
        community, as TDA methods and persistent homology are not areas I have
        a great deal of expertise in. The paper discussion and benchmarking
        analysis both seem sensible and well done. In general the methods they
        used are up to date, however the author's approach has enough
        limitations (detailed below) I think it should be revised before
        acceptance.

We thank the reviewer for a thorough and clear review. We address each of their
concerns below.

@@

We agree with the reviewer's point and have used `bench::bench_process_memory`
to collect memory benchmark data for all point cloud size, dimension, and
shape combinations. The resulting data has been visualized in Figure 7. Text
in the results and discussion sections has been adjusted to reflect the newly
added data. We thank the reviewer for the improvement and for providing a clear
method to correct our previous limitation.

        I think more discussion should be give to the big O `O()` runtime of the
        implementations. It is mentioned in the Fig. 3 legend that they all
        follow a power law, but at least from the data shown Ripser seems to
        have much lower coefficients, if it is indeed the same big O growth
        rate.

We agree with the reviewer's point. To address this, we have added a faceted
bar plot figure that illustrates the growth rate of each engine's calculation
function with various point cloud dimensions and shapes. Additionally, text in
the results and discussion sections has been edited/added to discuss the big O
complexity (memory and runtime).

### 2021-01-05 tdabench-paper/tdabench.Rmd
added big O stuff

```{r setup, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(broom))
```

@@

@@

\begin{figure}
  \centering
  \includegraphics[height=1.5in]{../Figures/Final_Figures/Intro_Alpha.png}
  \caption{\textbf{Basic visualization of the Alpha complex.}
  For a given $\alpha$, $\alpha$-radius balls are drawn around each point, and the union of the balls is taken.
  Then, an intersection between this union of $\alpha$-balls and the Vornoi diagram is taken.
  A connecting segment is drawn between points in adjacent Voronoi cells once the $\alpha$-ball fills out the Voronoi diagram.
  As $\alpha$ grows, more circles fill out the Voronoi cells.
  Once $\alpha$ is large enough, the Delaunay Triangulation is formed.}
\end{figure}

@@

@@

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/fig1.png}
  \caption{
    \textbf{Calculating persistent homology of a torus with three TDA libraries.}
    Median runtime (min to max, $n=10$ iterations per data point) for each TDA library (denoted by color) is plotted against point cloud size.
    Homological features up to 2 dimensions were calculated.
    Time complexity follows a power law for all three libraries (see GitHub repo for regression details).
    Although the libraries have similar runtimes for smaller point clouds, Dionysus has a clear disadvantage when the number of points exceeds 100.
    When the number of points exceeds 200, Ripser has a clear advantage over GUDHI, which maintains its advantage over Dionysus.
  }
\label{fig:tor}
\end{figure}

@@

\begin{widefigure}
  \centering
  \includegraphics[height=2in]{../Figures/Final_Figures/fig2.png}
  \caption{
    \textbf{Calculating persistent homology of round point clouds of varying dimensions with three TDA libraries.}
    Median runtime (min to max, $n=10$ iterations per data point) for each TDA library (denoted by color) is plotted against point cloud size and faceted by data dimension.
    Left panel compares library performance for a 2-dimensional circular point cloud; center panel for a 3-dimensional spherical point cloud; and right panel for a 4-dimensional hyperspherical point cloud.
    Maximum feature dimensions (one less than data dimension) were calculated in each case.
  }
\label{fig:cir}
\end{widefigure}

@@

# setup df
pl_exponents <- expand.grid(c("TDAstats", "GUDHI", "Dionysus"),
                            2:4,
                            c("circle", "annulus", "torus", "uniform"))

# return power law exponent from data
calc_pl_exp <- function(all_data, engine, dim, shape) {
  # only >= 100 points
  all_data <- all_data %>%
    filter(NumPoints >= 100 &
           Engine == engine &
           Dimension == dim &
           DataType == shape) %>%
    mutate(LogPts = log10(NumPoints),
           LogMem = log10(Memory))

@@

  return(tidy_mod$estimate[2])
}
```

In addition to runtime differences, the three Vietoris-Rips homology engines differ in memory use.
After crossing a threshold between 50 and 100 points, all three engines appeared to follow power law growth, with a linear trend on log-log plots (Figure \ref{fig:mem}).
However, for all combinations of point cloud dimension and shape, TDAstats used the least memory and Dionysus used the most, with TDAstats also growing with the smallest power law exponent as the number of points increased.
As an example, for 3-dimensional annulus point clouds, GUDHI memory use grew with a power law exponent of
`r calc_pl_exp(df_mem, "GUDHI", 3, "annulus") %>% signif(digits = 2)`,
Dionysus use grew with an exponent of
`r calc_pl_exp(df_mem, "Dionysus", 3, "annulus") %>% signif(digits = 2)`,
and TDAstats use grew with an exponent of
`r calc_pl_exp(df_mem, "TDAstats", 3, "annulus") %>% signif(digits = 2)`.
The entire table of exponents can be found in the Supplement (GitHub), but Figure \ref{fig:mem} clearly illustrates that TDAstats was, for all tested cases, the optimal Vietoris-Rips engine in the context of memory use.

@@

@@

Figure \ref{fig:fin} shows that at high point cloud sizes, GUDHI’s alpha complex outperforms Ripser. 
Theoretically, alpha complexes gain polynomial run time complexity as the number of points increases whereas Vietoris-Rips complexes gain exponential run time complexity \citep{roadmap}.
Specifically, alpha complexes are $O(n^{d/2})$ and Vietoris-Rips complexes are $O(2^n)$ where $n$ is the number of points on a point cloud and $d$ is the dimensions on the point cloud. 
For the conditions in our paper, Vietoris-Rips and alpha complexes both performed better than their theoretical maximums.
Vietoris-Rips complex calculations consistently had a polynomial growth for both runtime and boundary matrix size while alpha complexes had linear growth.

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/fig6.png}
  \caption{
    \textbf{Runtime comparison of persistent homology calculation between Ripser's Vietoris-Rips and GUDHI's alpha complex functionality.}
    Median runtime (min to max $n=10$ iterations per data point) for various 3-dimensional point cloud structures (facet) plotted against point cloud size for each library (color).
    Benchmarking was conducted on an annulus (top-left), a sphere (top-right), a torus (bottom-left), and a cube (bottom-right).
    Data was not collected for data dimensions greater than 3 due to computational limitations of calculating alpha complices.}
\label{fig:fin}
\end{figure}

@@

While Vietoris-Rips complexes can handle high-dimensional data well, calculation still significantly slows down when looking for higher dimension features. 
Boundary matrix size rapidly increases and can overwhelm available RAM even for small numbers of points. 
Looking for high-dimensional features in a large, high-dimensional point clouds remains a challenge. 
Methods to calculate persistent homology do exist for other simplicial complexes, such as the Delaunay complex and the Witness complex, but, to our knowledge, they are not currently implemented in CRAN packages.
Future challenges would be creating and implementing algorithms that reduce the computational complexity of higher dimensional topological feature calculations for R.

@@

The authors thank Emily Dolson, PhD for assistance with using a high-performance computing node for data collection.

>>>>

```{r setup, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(readr))
```

@@

\begin{figure}[b]
  \centering
  \includegraphics[width=4in]{../Figures/Final_Figures/Intro_Alpha.png}
  \caption{\textbf{Basic visualization of the Alpha complex.}
  For a given $\alpha$, $\alpha$-radius balls are drawn around each point, and the union of the balls is taken.
  Then, an intersection between this union of $\alpha$-balls and the Vornoi diagram is taken.
  A connecting segment is drawn between points in adjacent Voronoi cells once the $\alpha$-ball fills out the Voronoi diagram.
  As $\alpha$ grows, more circles fill out the Voronoi cells.
  Once $\alpha$ is large enough, the Delaunay Triangulation is formed.}
\end{figure}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/fig1.png}
  \caption{
    \textbf{Calculating persistent homology of a torus with three TDA libraries.}
    Median runtime (min to max, $n=10$ iterations per data point) for each TDA library (denoted by color) is plotted against point cloud size.
    Homological features up to 2 dimensions were calculated.
    Time complexity follows a power law for all three libraries (see GitHub repo for regression details).
    Although the libraries have similar runtimes for smaller point clouds, Dionysus has a clear disadvantage when the number of points exceeds 100.
    When the number of points exceeds 200, Ripser has a clear advantage over GUDHI, which maintains its advantage over Dionysus.
  }
\label{fig:tor}
\end{figure}

@@

\begin{widefigure}[bt]
  \centering
  \includegraphics[height=2in]{../Figures/Final_Figures/fig2.png}
  \caption{
    \textbf{Calculating persistent homology of round point clouds of varying dimensions with three TDA libraries.}
    Median runtime (min to max, $n=10$ iterations per data point) for each TDA library (denoted by color) is plotted against point cloud size and faceted by data dimension.
    Left panel compares library performance for a 2-dimensional circular point cloud; center panel for a 3-dimensional spherical point cloud; and right panel for a 4-dimensional hyperspherical point cloud.
    Maximum feature dimensions (one less than data dimension) were calculated in each case.
  }
\label{fig:cir}
\end{widefigure}

@@

@@

@@

# setup df
pl_exponents <- expand.grid(c("TDAstats", "GUDHI", "Dionysus"),
                            2:4,
                            c("circle", "annulus", "torus", "uniform"),
                            stringsAsFactors = FALSE) %>%
  filter(Var3!="torus" | Var2 == 3) %>%
  rename(Engine = Var1,
         Dimension = Var2,
         Shape = Var3)

# return power law exponent from data
calc_pl_exp <- function(all_data, engine, dim, shape) {
  # only >= 100 points
  all_data <- all_data %>%
    filter(NumPoints >= 25 &
           Engine == engine &
           Dimension == dim &
           DataType == shape) %>%
    mutate(LogPts = log10(NumPoints),
           LogMem = log10(Memory))

@@

  return(tidy_mod$estimate[2])
}

calc_pl_exp_summ <- function(size, value) {
  log_pts <- log10(size)
  log_mem <- log10(value)

  lm_mod <- lm(log_mem ~ log_pts)
  tidy_mod <- tidy(lm_mod)

  return(tidy_mod$estimate[2])
}

# add calculated values to pl_exponents as an additional column
exp_vec <- vapply(X = seq_len(nrow(pl_exponents)),
                      FUN.VALUE = numeric(1),
                      FUN = function(row_num) {
                        calc_pl_exp(df_mem,
                                    pl_exponents$Engine[row_num],
                                    pl_exponents$Dimension[row_num],
                                    pl_exponents$Shape[row_num])
                      }
                  )
pl_exponents <- pl_exponents %>%
  mutate(Exp = exp_vec)
```

In addition to runtime differences, the three Vietoris-Rips homology engines differ in memory use.
All three engines appeared to follow power law growth, with a linear trend on log-log plots (Figure \ref{fig:mem}).
However, for nearly all combinations of point cloud dimension and shape, TDAstats used the least memory and Dionysus used the most, with TDAstats also growing with the smallest power law exponent as the number of points increased for most point clouds.
For most point clouds, runtime and memory complexity for TDAstats (Ripser) grew with a power function at least one degree less than the other engines (Figure \ref{fig:bigO}).

@@

```{r bigoh, message=FALSE, warning=FALSE,echo=FALSE, fig.height=3, fig.cap="\\label{fig:bigO}\\textbf{Big-O exponents for runtime and memory complexity as point cloud size varies.} In the majority of cases, Ripser (and consequently, TDAstats) has the lowest exponent, indicating slowest growth in complexity as point cloud size increases. Due to shape constraints, torus only has data available in the third dimension."}
# collect data for runtime
get_runtime_data <- function(shape) {
  filepath <- paste0("../Cluster_data/",
                     shape,
                     ".csv")

  # read and transform data
  read_csv(filepath) %>%
    select(-c(X1, V1)) %>%
    rename(Shape = V2,
           Dimension = V3,
           Size = V4,
           Calc_Dim = V5,
           Engine = V6) %>%
    filter(Calc_Dim == Dimension - 1) %>%
    filter(Engine != "GUDHIalpha") %>%
    pivot_longer(cols = paste0("V", 7:16)) %>%
    select(-c(name, Calc_Dim)) %>%
    group_by(Shape, Dimension, Size, Engine) %>%
    summarise(Value = median(value)) %>%
    mutate(Engine = gsub(x = Engine,
                         pattern = "stats",
                         replacement = "TDAstats")) %>%
    rename(DataType = Shape)
}

# collect data
annulus <- get_runtime_data("annulus")
circle <- get_runtime_data("circle")
torus <- get_runtime_data("torus")
uniform <- get_runtime_data("uniform")

runtime_data <- rbind(annulus, circle, torus, uniform) %>%
  filter(Dimension %in% 2:4) %>%
  group_by(Engine, Dimension, DataType) %>%
  summarise(Exp = calc_pl_exp_summ(Size, Value)) %>%
  mutate(Type = "runtime")

memory_data <- pl_exponents %>%
  rename(DataType = Shape) %>%
  mutate(Type = "memory")

combo_data <- rbind(runtime_data, memory_data) %>%
  mutate(Engine = gsub(Engine, pattern = "TDAstats", replacement = "Ripser"))

# plot big O exponents for memory
ggplot(combo_data, aes(x = Dimension, y = Exp, fill = Engine)) +
  facet_grid(Type ~ DataType) +
  geom_bar(stat = "identity", position = "dodge") +
  ylab("Big-O exponent") +
  scale_y_continuous() +
  scale_fill_manual(values = c("red", "blue", "brown")) +
  theme_bw()
```

@@

Figure \ref{fig:fin} shows that at high point cloud sizes, GUDHI’s alpha complex outperforms Ripser. 
Theoretically, alpha complexes gain polynomial run time complexity as the number of points increases whereas Vietoris-Rips complexes gain exponential run time complexity \citep{roadmap}.
Specifically, alpha complexes are $O(n^{d/2})$ and Vietoris-Rips complexes are $O(2^n)$ where $n$ is the number of points on a point cloud and $d$ is the dimensions on the point cloud. 
For the conditions in our paper, Vietoris-Rips and alpha complexes both performed better than their theoretical maximums.
Vietoris-Rips complex calculations consistently had a polynomial growth for both runtime and memory, while alpha complexes had linear runtime growth.

\begin{figure}
  \centering
  \includegraphics[width=6in]{../Figures/Final_Figures/fig6.png}
  \caption{
    \textbf{Runtime comparison of persistent homology calculation between Ripser's Vietoris-Rips and GUDHI's alpha complex functionality.}
    Median runtime (min to max $n=10$ iterations per data point) for various 3-dimensional point cloud structures (facet) plotted against point cloud size for each library (color).
    Benchmarking was conducted on an annulus (top-left), a sphere (top-right), a torus (bottom-left), and a cube (bottom-right).
    Data was not collected for data dimensions greater than 3 due to computational limitations of calculating alpha complices.}
\label{fig:fin}
\end{figure}

@@

While Vietoris-Rips complexes can handle high-dimensional data well, calculation still significantly slows down when looking for higher dimension features.
This is evidenced by the big-O polynomial growth for runtime and memory that have degree less than 4 for most 2-dimensional point clouds, but degree between 4 and 6 for most 4-dimensional point clouds (Figure \ref{fig:bigO}); even higher degree complexities should be expected as point cloud dimension increases.
Thus, finding high-dimensional topological features in high-dimensional point clouds remains a challenge. 
Methods to calculate persistent homology do exist for other simplicial complexes, such as the Delaunay complex and the Witness complex, but, to our knowledge, they are not currently implemented in CRAN packages.
Future challenges would be creating and implementing algorithms that reduce the computational complexity of higher dimensional topological feature calculations for R.

@@

The authors thank Emily Dolson, PhD for assistance with using a high-performance computing node for data collection.
JGS was supported by NIH grant R37 CA244613.
